---
title: "Practical Machine Learnng 
subtitle: Course Project"
author: "Diane Wenke"
date: "November 19, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Background

- Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. 
- These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 
- One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
- In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Data

The training data for this project are available here:

- https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

- https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: 

- http://groupware.les.inf.puc-rio.br/har

If you use the document created for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

# Project Goal
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

# Project Submission

Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

# Project Quiz
Apply your machine learning algorithm to the 20 test cases available in the test data above and submit your predictions in appropriate format to the separate Course Project Prediction Quiz for automated grading.

# Project Packages 
(Required to generate the results)

- To install packages (e.g. rattle package) in R, run this code from the console command line: install.packages("rattle")
- For this project the following libraries were used:

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```


```{r}
set.seed(12345)
```

# Get Train & Test Data Sets
- The training data set can be found on the following URL:
```{r}
trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```

- The testing data set can be found on the followng URL:
```{r}
testURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

# Load Train & Test Data Sets
- Load the training data set
```{r}
training <- read.csv(url(trainURL), na.strings=c("NA","#DIV/0!",""))
```

- Load the testing data set
```{r}
testing <- read.csv(url(testURL), na.strings=c("NA","#DIV/0!",""))
```

# Partition the Training Data Set
- Need to partition the training data set into two data sets: 60% for myTraining and 40% for myTesting
```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ];
myTesting <- training[-inTrain, ]
dim(myTraining);
dim(myTesting)
```

# Data Cleansing
Step1:
```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics = TRUE)
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")

myTraining <- myTraining[!myNZVvars]
```
# Check the number of observations
```{r}
dim(myTraining)
```
Step2:

- Delete the first column "ID" so it will not negatively impact the Machine Learning Algorithms
```{r}
myTraining <- myTraining[c(-1)]
```
# Check the number of observations
```{r}
dim(myTraining)
```
Step3:

- Omit variables with more than 60% thresholds of NA's

```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if nº NAs > 60% of total observations
		for(j in 1:length(trainingV3)) {
			if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
				trainingV3 <- trainingV3[ , -j] #Remove that column
			}	
		} 
	}
}
```
# Check the number of observations
```{r}
dim(trainingV3)
```
# Assign data set name back to myTraining
```{r}
myTraining <- trainingV3
```

# Repeat steps 1-3 with the data sets - 
# myTesting and testing: 
Steps 1-3:
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]
```
# Check the number of observations
```{r}
dim(myTesting)
```
# Check the number of observations
```{r}
dim(testing)
```

- Note: The last column - problem_id - which is not equal to training sets, was also removed

- In order to ensure proper functioning of Decision Trees and especially RandomForest 
Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.
```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
		if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
			class(testing[j]) <- class(myTraining[i])
		}      
	}      
}
```
- To ensure the Coertion worked properly, use this step:
```{r}
testing <- rbind(myTraining[2, -58] , testing) 
```
- Removing row 2 as it does not add value
```{r}
testing <- testing[-1,]
```

# Using Machine Learning Algorithms for Prediction:
# Decision Tree
```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```
- Note: to view the decision tree with fancy run this command:
```{r}
fancyRpartPlot(modFitA1)
```

# Predicting Step:

```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

# Test Results: Using Confusion Matrix
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

# Overall Statistics

                Accuracy : 0.8789          
                  95% CI : (0.8715, 0.8861)
     No Information Rate : 0.2845
      P-Value [Acc > NIR] : < 2.2e-16
             
# Using Machine Learning Algorithms for prediction: 
# Random Forests

```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```
# Predicting:
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```
# Using Confusion Matrix to test results:

```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```
# Overall Statistics
                                          
                Accuracy : 0.9986          
                  95% CI : (0.9975, 0.9993)
     No Information Rate : 0.2845          
     P-Value [Acc > NIR] : < 2.2e-16       
                                          
                   Kappa : 0.9982          
  Mcnemar's Test P-Value : NA 
  
# Random Forests yielded better Results

# Generating Files with Answers to submit:

For Random Forests:
```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```
# Generate files with Predictions to submit:
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```
# Ploting Result Summary

```{r, echo=FALSE}
plot(predictionsB2)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
